---
title: "6_batch_correction"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
---


6_batch_correction.rmd Preliminary Checks and Cleaning

     a) Clean the abundance table: remove zero-abundance taxa and standardize column names.
     b) Optional Batch Effect Correction. If necessary, batch effect correction is performed at this stage using the ConQuR package.

Conditional Quantile Regression (ConQuR)  - https://www.nature.com/articles/s41467-022-33071-9
package ConQuR -  ConQuR : https://github.com/wdl2459/ConQuR
Vignette: https://wdl2459.github.io/ConQuR/ConQuR.Vignette.html


```{r}
library(dplyr)
library(tidyverse)
library(vegan)
library(coin)
library(pheatmap)


# devtools::install_github("wdl2459/ConQuR")
library(ConQuR)
library(doParallel)
```

####################################################################################################
# 0 LOAD DATA
####################################################################################################

```{r}
# kraken result:
species_kraken <- read.table(paste0(dir_data,"bracken_abundance_species_mpa.txt"), check.names = FALSE) 

# Modify or standardize batch ID for downstream processing . Colname: ("ID_samples","batch").

id_mostres <- read.table("_____.txt")

# Optional: include covariates if applicable
covariates <- read.table("_____.txt")
```


####################################################################################################
# a) Clean the abundance table: remove zero-abundance taxa and standardize column names.
####################################################################################################

## 1) clean_table

```{r}
# colnames modification

colnames(species_kraken) <-gsub("C_nohuman_Q20.report_bracken.txt", "", colnames(species_kraken))
colnames(species_kraken) <-gsub("B_nohuman_Q20.report_bracken.txt", "", colnames(species_kraken))
colnames(species_kraken) <-gsub("C_nohuman_Q20report_bracken.txt", "", colnames(species_kraken))
colnames(species_kraken) <-gsub("B_nohuman_Q20report_bracken.txt", "", colnames(species_kraken))

```

## 2) all 0 sp - NA

```{r}

table(rowSums(species_kraken) == 0)
species_kraken <- species_kraken[!rowSums(species_kraken) == 0,]

```




####################################################################################################
# b) Optional Batch Effect Correction. If necessary, batch effect correction is performed at this stage using the ConQuR package.
####################################################################################################

##   1) Data preparation
#########################

```{r}
# sp extraction 
species_kraken$species <- sub(".*s__([^|]+).*", "s__\\1", rownames(species_kraken))

# Save lineage and species information in a separate DataFrame to enable lineage retrieval at a later stage.
kraken_llinatge_especie <- data.frame(Lineage = rownames(species_kraken), species = species_kraken$species)

# Remove lineage information from the main DataFrame
rownames(species_kraken) <- species_kraken$species
species_kraken$species <- NULL
```


```{r}
# TAX table

#samples- orw / taxa - col
tax_table <- as.data.frame(t(species_kraken))
```

```{r}
# Order the taxonomy table and sample IDs consistently based on sample identifiers
identical(id_mostres$V1 , rownames(tax_table))
id_mostres <- id_mostres[match(rownames(tax_table), id_mostres$V1),]
identical(id_mostres$V1, rownames(tax_table))
batch_id <- as.factor(id_mostres$batch)
```

```{r}
# Order covariates table and sample IDs consistently based on sample identifiers
identical(covariates$id_icobiobanc , rownames(tax_table))
table(covariates$id_icobiobanc %in% rownames(tax_table) )
covariates <- covariates[match(rownames(tax_table) ,covariates$id_icobiobanc ),]
```


##   2 APLICACIO ConQuR 
#########################


```{r}
options(warn=-1)
taxa_corrected1 = ConQuR(tax_tab=tax_table, batchid=batch_id, covariates=covariates, batch_ref="batch1",num_core = 20  )

saveRDS(taxa_corrected1, "taxa_batch_corrected.rds") # -> data to use

```



#########################
##   3. VALIDATION STEPS
#########################

```{r}
# Convert tax_table and corrected table to data frames
df1 <- as.data.frame(tax_table) 
df2 <- as.data.frame(taxa_corrected1)

# Check if all zeros in df1 are also zeros in df2
zero_comparison <- (df1 == 0) & (df2 == 0)
if(all((df1 == 0) == zero_comparison)) {
  print("All zeros in df1 are preserved in df2.")
} else {
  print("Discrepancy found: not all zeros in df1 are zeros in df2.")
  discrepant_positions <- which((df1 == 0) & (df2 != 0), arr.ind = TRUE)
  num_discrepancies <- nrow(discrepant_positions)
  print(paste("Number of zero mismatches:", num_discrepancies))
}

# Force df2 to have zero where df1 has zero
df2[df1 == 0 & df2 != 0] <- 0

# Recheck after correction
zero_comparison <- (df1 == 0) & (df2 == 0)
if(all((df1 == 0) == zero_comparison)) {
  print("Zero correction successful: df2 now matches df1.")
} else {
  print("Discrepancies remain after correction.")
}

# Save corrected taxa table
saveRDS(taxa_corrected1, "0.2_taxa_corrected.rds")
```




##   4. SAVE FORMATTED DATA
#########################


```{r}
# Transpose for downstream analysis (samples as rows)
df2 <- t(df2)

# Check matching IDs and update rownames with full lineage
table(rownames(df2) %in% kraken_llinatge_especie$species)
identical(rownames(df2), kraken_llinatge_especie$species)

rownames(df2) <- kraken_llinatge_especie$Lineage

# Save final version
saveRDS(df2, "taxa_batch_corrected_0.rds")
```



