---
title: "7_taxonomic_data_preparation.rmd"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
---


**ATENTION!** If batch effect correction is not applied: clean the abundance table by removing zero-abundance taxa and standardizing column names. 
See: scripts/1_Metagenomics_pipeline/6_batch_correction.rmd - 
          a) Clean the abundance table: remove zero-abundance taxa and standardize column names.



- Normalize species-level abundances by genome length to obtain comparable abundance measures across taxa.

Steps:
    1. Load genome length data and manifest.txt file
    2. Harmonize species names between genome_length and Kraken species data
    3. Normalize abundance values by genome length
  

```{r}
library(dplyr)
library(tidyverse)
library(vegan)
library(coin)
library(pheatmap)


```

####################################################################################################
# 0 LOAD DATA
####################################################################################################


## kraken - batch effect corrected result

```{r}
# kraken - batch effect corrected result:
species_kraken <- readRDS("taxa_batch_corrected_0.rds")
species_kraken <- as.data.frame(species_kraken)

```
## load genome_lenght

The Kraken database should provide a table listing species and their corresponding genome lengths.

Note: The format of species names (lineages) in the Kraken/Bracken output and in the genome length reference file might differ.  
A normalization or mapping step may be necessary to ensure consistent matching between both datasets (not provided).

Equiparació realitzada amb script *equiparacio_kraken.Rmd*. Veure: */mnt/typhon/data/references/metagenomics_dynamic/uhgg/2.0.2/*


```{r}
# Load genome length data
# - Sort species in ascending order
# - Convert genome lengths from base pairs to megabases (MB)

genome_length <- read.table("_____") %>%
    arrange(species) %>%                     # Sort by species name
    mutate(length = Length / 1e6)            # Convert base pairs to megabases (MB)

# Ensure the 'length' column is of numeric type
genome_length$length <- as.double(genome_length$length)

# Remove duplicated lineages, keeping the first occurrence
genome_length <- distinct(genome_length, Lineage, .keep_all = TRUE)

```

####################################################################################################
# 2 Data preparation
####################################################################################################


```{r}
# Transform Kraken results dataframe into a matrix:
#   - Columns represent species
#   - Rows represent patient/sample IDs

# Extract species names from the taxonomic lineage
species_kraken$species <- sub(".*s__([^|]+).*", "s__\\1", rownames(species_kraken))

# Save full lineage and extracted species into a separate dataframe for later reference
kraken_lineage_species <- data.frame(Lineage = rownames(species_kraken), species = species_kraken$species)

# Remove taxonomic lineage from the main dataframe and set species as row names
rownames(species_kraken) <- species_kraken$species
species_kraken$species <- NULL

```

Prepare genome_length and kraken_results tables to contain the same species and in the same order for comparison.

```{r}
# Select only species from genome_length that are present in species_kraken
table(row.names(species_kraken) %in% genome_length$species)

genome_length <- genome_length[genome_length$species %in% row.names(species_kraken), ]
table(genome_length$species %in% row.names(species_kraken))

# Sort both data frames by species to ensure identical order
genome_length <- genome_length[order(genome_length$species), ]
species_kraken <- species_kraken[order(row.names(species_kraken)), ]

# Confirm that both lists are now aligned by species
identical(row.names(species_kraken), genome_length$species)

```

####################################################################################################
# 3 Normalization
####################################################################################################


```{r}
# Avoid scientific notation in output (e.g., 1e+06)
options(scipen = 999)

# Normalize species abundances by genome length (in megabases)
species_all_normalized <- as.data.frame(species_kraken / genome_length$length)

```

```{r}
table(rowSums(species_all_normalized) == 0)

```



```{r}
# Rename rownames using full taxonomic lineage

# Add species as a column to match lineages
species_all_normalized$species <- rownames(species_all_normalized)

# Confirm matching species between normalized table and lineage metadata
table(species_all_normalized$species %in% kraken_llinatge_especie$species)

# Join to retrieve full lineage
species_all_normalized <- left_join(species_all_normalized, kraken_llinatge_especie, by = "species")

# Replace rownames with lineage info
rownames(species_all_normalized) <- species_all_normalized$Lineage

# Remove helper columns
species_all_normalized$Lineage <- NULL
species_all_normalized$species <- NULL

# View final normalized table
species_all_normalized

# Save the batch-effect-corrected and genome-normalized table
save(species_all_normalized, file = "1_shotgun_species_BEC_normalized.RData", compress = "xz")
write.table(species_all_normalized, "1_shotgun_species_BEC_normalized.txt", col.names = TRUE, sep = "\t")

```



